---
title: "INFO 534 Homework 1"
author: "Rachel Sagers"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Q1(a)**

The dimension of the resulting matrix C from multiplying A times B will be 4 by 4. Matrix A is *mxn* (4,3), and matrix B is *nxo* (3,4), so the resulting matrix C will be *mxo* (4,4). 

**Q1(b)**

The value of C~2,2~ can be computed by multiplying the vector of row 2 of matrix A, by the vector of column 2 of matrix B and summing the products. The value of C~2,2~ will thus be *5(3) + 1(1) + 2(1) = 18.*

**Q1(c)**

```{r matrices}
# produce matrix A
matrixA <- matrix(c(4,5,-1,2,2,1,5,-3,2,2,1,3), nrow=4, ncol=3)

# produce matrix B
matrixB <- matrix(c(2,4,1,3,1,1,1,-3,-1,3,1,1), nrow=3, ncol=4)

# matrix multiplication
matrixC <- matrixA %*% matrixB

# resulting matrix C
print(matrixC)
```

**Q2**

```{r while}
# while loop
v <- 10
while(15 - v > 0.1)
{
  v <- v + runif(n=1, min=0, max=1)
  print(v)
}
```

**Q3(a)**

```{r histogram}
# generate 5000 random values of x and create histogram
x <- runif(n=5000, min=0, max=1000)
hist(x)
```

**Q3(b)**

``` {r linear model}
# linear model function
output = function(a,x){
y = -1 + a*x
return(y)
}

# generate y values from x values using function
y <- output(x=x, a=0.2)

# plot of x vs y
plot(x,y)
```

**Q4(a)**

```{r log likelihood}
# the range of the model parameter value to be considered
a.values = seq(-2,2,0.1)

# log likelihood function
logL = function(a, x, y){
  sigma=1
  500*log(1/(sqrt(2*pi)*sigma)) - 1/(sigma^2)*(1/2)*sum((a*x - y)^2)
}

# create an empty vector to store output of for loop 
logoutput <- numeric()

# for loop to input all a values in function logL
for(a in a.values) {
  output <- logL(a=a, x=x, y=y)
  logoutput <- c(logoutput, output)
}

# plot of a values vs logL function output
plot(a.values, logoutput)
```

By visually inspecting the resulting graph, I would estimate that the value of *a* that corresponds to the highest value of the log-likelihood is about **0.4**.

**Q4(b)**

```{r optimmle}
optim(par=0, logL, x=x, y=y,method = "L-BFGS-B",control = list(fnscale=-1))
```

If the function argument *fnscale* is set to -1, it will lead to a maximization of the defined log-likelihood. 

**Q4(c)**

```{r mle}
mleresult <- optim(par=0.5, logL, x=x, y=y,method = "L-BFGS-B",control = list(fnscale=-1))
print(mleresult$value)
```
Yes, the value of the MLE *-1080.794* with the input parameter *0.5* is consistent with the answer in question 4(a).